---
title: "Premier League Soccer"
description: |
  Analyzing 2021-2022 Premier League season match data
author: Charlotte Imbert
date: September 13, 2024
format: html
---

## The Data

The dataset for this analysis was accessed via <https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-04-04> . The data is from the [Premier League Match Data 2021-2022](#0) via [Evan Gower](#0) on Kaggle. The dataset contains information about soccer matches in the 2021-2022 English Premier League season, including the date, referee, home and away team, full-time home and away goals, home and away fouls, in addition other information for each match.

```{r message=FALSE}
library(tidytuesdayR)
library(tidyverse)
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-04/soccer21-22.csv')
```

## Fouls and Relegation Status

I am interested in identifying trends between the total number of fouls a team commits and their relegation status. The bottom three teams at the end of each season are relegated to the Championship. Relegation status for this analysis is based on the season-end Premier League standings in 2022, obtained from the Premier League website at <https://www.premierleague.com/tables?co=1&se=418&ha=-1>.

Firstly, listing teams in descending order of total fouls at home:

```{r}
soccer_summary<-soccer |>
  group_by(HomeTeam) |>
  summarize(totalfouls=sum(HF)) |>
  arrange(desc(totalfouls))
head(soccer_summary)
```

Now, plotting this relationship and highlighting the teams that were relegated at the end of the season, in navy blue:

```{r}
plot<-ggplot(soccer_summary, aes(x=reorder(HomeTeam, totalfouls), y=totalfouls)) +
  geom_bar(stat = "identity", aes(fill = ifelse(HomeTeam %in% c("Burnley", "Norwich", "Watford"), "Relegated", "Safe"))) +
  scale_fill_manual(values = c("Relegated" = "navyblue", "Safe" = "lightblue")) +
  theme_minimal() +
  labs(title="Premier League Fouls by Team, 2021-2022",
       subtitle="Home Games",
       y="Total Number of Fouls",
       x="Team",
       fill="Relegation Status")
plot + coord_flip()
```

Horizontal bar plot showing the total number of fouls in home games for each team in the 2021-2022 Premier League season. Teams that were relegated at the end of the season are colored in navy blue.

Is this relationship different when teams are playing away from home?

```{r}
soccer_summary2<- soccer |>
  group_by(AwayTeam) |>
  summarize(totalfouls_a=sum(AF)) |>
  arrange(desc(totalfouls_a))
head(soccer_summary2)
```

```{r}
plot2<-ggplot(soccer_summary2, aes(x=reorder(AwayTeam, totalfouls_a), y=totalfouls_a)) +
  geom_bar(stat = "identity", aes(fill = ifelse(AwayTeam %in% c("Burnley", "Norwich", "Watford"), "Relegated", "Safe"))) +
  scale_fill_manual(values = c("Relegated" = "red", "Safe" = "lightpink")) +
  theme_minimal() +
  labs(title="Premier League Fouls by Team, 2021-2022",
       subtitle="Away Games",
       y="Total Number of Fouls",
       x="Team",
       fill="Relegation Status")
plot2 + coord_flip()
```

Horizontal bar plot showing the total number of fouls in away games for each team in the 2021-2022 Premier League season. Teams that were relegated at the end of the season are colored in red.

## Conclusion

Based on these plots, a team's total number of fouls does not seem to influence their relegation status. For example, Norwich had one of the lowest number of fouls away from home but still faced relegation.

We can also see that home status does not really influence the total number of fouls committed by a team: Leeds and Manchester City are top and bottom, respectively, for the number of fouls both at home and away. The majority of the teams in the league committed a similar total number of fouls at home and away. Manchester United are an exception as they committed the second highest number of fouls away, compared to the fifth lowest at home.
